A Large and Variable Leading Tail of Helium in a Sub-Saturn Undergoing Runaway Inflation

### 2023-06-26 paper/paper1/ms.tex
Page 2 reference additions

\begin{abstract}
    Atmospheric escape and its associated mass loss shape the fate of exoplanets, with statistical evidence for photoevaporation imprinted across the mass-radius-insolation distribution. The Habitable Zone Planet Finder (HPF) Spectrograph Exospheres program aims to expand the search across a diverse range of planet and host star properties via the 10830 \AA~ Helium triplet, a sensitive probe of active atmospheric escape. Here we present a detection of 10\% absorption depth during transit spectroscopy of the highly irradiated, low-gravity, inflated hot Saturn HAT-P-67b. The 13.8 hours of on-sky integration time over 39 nights provides enough out-of-transit phase coverage as to reveal comparably deep absorption preceding the transit, revealing a large leading tail. This configuration can be understood as the escaping material overflowing its small Roche lobe, and advecting most of the gas into the stellar—and not planetary---rest frame, consistent with the Doppler velocity structure seen in the Helium line profiles. The prominent leading tail serves as direct evidence for dayside mass loss with a strong day-/night- side asymmetry.   A combination of Keplerian shear and stellar wind confinement control the extent and optical depth of the tail.  We see some transit-to-transit variability in the line profile, consistent with the interplay of stellar and planetary wind.  We employ 1D Parker wind models to estimate the mass loss rate, finding values on the order of $2\times10^{13}$ g/s, with large uncertainties owing to the unknown XUV flux of the F host star.  The large mass loss in HAT-P-67~b represents a valuable example of an inflated hot Saturn, a class of planets recently identified to be rare as their atmospheres are predicted to evaporate quickly.  We explore two physical mechanisms for runaway evaporation: Ohmic Dissipation and XUV irradiation, slightly favoring the latter.   
\end{abstract}

@@

Inflated hot Saturns stand out as an especially extreme category that should exhibit mass loss.  This category is defined as having masses at or somewhat less than Saturn's ($M_p \lesssim 0.3 M_\mathrm{Jup}$), with equilibrium temperatures high enough to cause radius inflation($T_\mathrm{eq}>1000\;$K).  Their low gravitational potentials should let go of their atmospheres more readily than their steadfast hot Jupiter counterparts.  Their lower gravity yields larger atmospheric scale heights, making them easier to detect in transmission spectroscopy.  And their large transit depths and short periods should make them readily detectable in large numbers, like hot Jupiters.

But inflated hot Saturns are rare \citep{2018AJ....155..214T}.  The cause for their underabundance remains an open question, with two leading candidate explanations.  Tidal migration mechanisms may be mass-dependent in such a way as to proceed efficiently for Jupiter-mass planets, but inefficiently for the lower-mass Saturns \citep{2018AJ....155..214T,2018ARA&A..56..175D}.  In this scenario, sub-Saturn mass planets never make it to the close-in orbital separations that would lead to the conditions needed for inflation in the first place.

@@

A runaway inflation scenario predicts an inevitable and profound mass loss rate for inflated hot Saturns.  As the planet's atmosphere overflows its Roche lobe at an ever-increasing pace, instantaneous mass loss rates may exceed $\dot{M}>10^{13}$ g/s, over 10$\times$ larger than those previously seen in atmospheric escape measurements to date \citep{2022arXiv221116243D}.  Inflated hot Saturns, therefore, make excellent targets for direct measurement of atmospheric escape.

Large mass loss alone does not guarantee detectability.  The ability to detect even immense mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2018ApJ...855L..11O,2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own observational limitations.  Ly$\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ 3470 b} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  An observational picture of mass loss in hot Saturns appears to be lacking, at least in part due to their intrinsic rarity.

@@

\subsection{Ohmic Dissipation-driven mass loss}
The Ohmic dissipation mechanism made two key observable predictions.  Anomalous heating efficiency, $\epsilon$, should initially increase as a function of equilibrium temperature, then degrade as irradiation exceeds equilibrium temperatures of about 2000 K.  Second, hot Saturns ($\lesssim0.5 M_\mathrm{Jup}$) should undergo runaway evaporation, whereas hot Jupiters ($\gtrsim 1 M_\mathrm{Jup}$) should reach stable equilibrium---albeit inflated---radii after Gyr timescales.  Both of these outcomes predated data that could validate them, and these outcomes differ from the behavior of other heating mechanisms, such as photoevaporation or tides.

@@

\noindent yielding an incredibly short $<$5 Myr timescale for \hatpb, assuming a typical $\epsilon\sim0.01$.  The order of magnitude of this inflation timescale is so fleetingly short that---according to this scenario---\hatpb~must be in the runaway stage of inflation, rapidly losing mass, and growing in surface area to fuel a positive feedback loop.  Under this interpretation, \hatpb~ would represent an example of a new category of planet system that is doomed to evaporate entirely due to the Ohmic Dissipation mechanism, as predicted by \citet{2011ApJ...738....1B}.  At Roche lobe overflow, a 5 Myr inflation timescale may imply an instantaneous $\dot{M}_\mathrm{infl}>10^{3}$ M$_\oplus$/Gyr.

@@

The second panel from the top illustrates a countervailing effect: the increase in flux actually triggers a decrease in anomalous heating efficiency, $\epsilon$. The greater stellar energy couples into the planet less effectively than the weaker radiation did, roughly balancing out. Together the effects nearly cancel when computing the inflationary lifetimes, yielding nearly identical curves in the bottom panel: a secularly evolving main sequence history gives almost the same inflation timescale as a rapidly increasing subgiant.  It is important to emphasize that the anomalous heating efficiency found by \citet{2018AJ....155..214T} is agnostic to what the actual heating mechanism is: either XUV Irradiation or Ohmic Dissipation both have to obey the trend of $\hat{\epsilon}_G(T_\mathrm{eq})$.

@@

We report in-transit line profile variability, which we attribute to the delicate interplay of planetary and stellar winds.  We identify several avenues for future work, including additional monitoring of the line profile variability to probe the stellar-and-planetary wind interaction.  The large signal should be perceptible in other spectral tracers, such as metal lines in the UV.  We offer a list other planets that may be likely to exhibit mass loss under the Ohmic Dissipation and XUV irradiation scenarios.

>>>>

\begin{abstract}
    Atmospheric escape and its associated mass loss shape the fate of exoplanets, with statistical evidence for photoevaporation imprinted across the mass-radius-insolation distribution. The Habitable Zone Planet Finder (HPF) Spectrograph Exospheres program aims to expand the search across a diverse range of planet and host star properties via the 10830 \AA~ Helium triplet, a sensitive probe of active atmospheric escape. Here we present a detection of 10\% absorption depth during transit spectroscopy of the highly irradiated, low-gravity, inflated hot Saturn HAT-P-67b. The 13.8 hours of on-sky integration time over 39 nights provides enough out-of-transit phase coverage as to reveal comparably deep absorption preceding the transit, revealing a large leading tail. This configuration can be understood as the escaping material overflowing its small Roche lobe, and advecting most of the gas into the stellar—and not planetary---rest frame, consistent with the Doppler velocity structure seen in the Helium line profiles. The prominent leading tail serves as direct evidence for dayside mass loss with a strong day-/night- side asymmetry.   A combination of Keplerian shear and stellar wind confinement control the extent and optical depth of the tail.  We see some transit-to-transit variability in the line profile, consistent with the interplay of stellar and planetary wind.  We employ 1D Parker wind models to estimate the mass loss rate, finding values on the order of $2\times10^{13}$ g/s, with large uncertainties owing to the unknown XUV flux of the F host star.  The large mass loss in HAT-P-67~b represents a valuable example of an inflated hot Saturn, a class of planets recently identified to be rare as their atmospheres are predicted to evaporate quickly.  We explore two physical mechanisms for runaway evaporation: Ohmic dissipation and XUV irradiation, slightly favoring the latter.   
\end{abstract}

@@

Inflated hot Saturns stand out as an especially extreme category that should exhibit mass loss.  This category is defined as having masses comparable to Saturn's ($M_p \sim 0.3 M_\mathrm{Jup}$), with equilibrium temperatures high enough to expect radius inflation($T_\mathrm{eq}>1000\;$K).  Their low gravitational potentials should let go of their atmospheres more readily than their steadfast hot Jupiter counterparts.  Their lower gravity yields larger atmospheric scale heights, making them easier to detect in transmission spectroscopy.  And their large transit depths and short periods should make them readily detectable in large numbers, like hot Jupiters.

But inflated hot Saturns are rare \citep{2018AJ....155..214T}.  The cause for their underabundance remains an open question, with at least two conceivable explanations.  Tidal migration mechanisms---either high-eccentricity or disk-based---could hypothetically proceed in a mass-dependent manner, efficiently for Jupiter-mass planets, but inefficiently for the lower-mass Saturns \citep{2018AJ....155..214T,2018ARA&A..56..175D}.  In this scenario, sub-Saturn mass planets never make it to the close-in orbital separations that would lead to the conditions needed for inflation in the first place.

@@

A runaway inflation scenario predicts an inevitable and profound mass loss rate for inflated hot Saturns.  As the planet's atmosphere overflows its Roche lobe at an ever-increasing pace, instantaneous mass loss rates may exceed $\dot{M}>10^{13}$ g/s, over 10$\times$ larger than those previously seen in atmospheric escape measurements to date \citep{2022arXiv221116243D}.  Inflated hot Saturns, therefore, make excellent targets for direct measurement of atmospheric escape and to test the underlying mechanisms of mass loss.

Large mass loss alone does not guarantee detectability.  The ability to detect even immense mass loss hinges on its observability in spectral tracers.  Ly$\alpha$, \ion{He}{1} 10833 \AA, and H$\alpha$ have emerged as the most amenable to detection \citep{2000ApJ...537..916S,2003Natur.422..143V,2012ApJ...751...86J,2018NatAs...2..714Y,2018ApJ...855L..11O,2018Natur.557...68S, 2022arXiv221116243D,2023MNRAS.518.4357O}, but each of these has its own observational limitations.  Ly$\alpha$ can suffer from interstellar medium \ion{H}{1} absorption censoring its low-velocity line core, for example.  Here we focus on \ion{He}{1} 10833 \AA, which offers some advantages.  In particular, metastable Helium's ability to be observed at high spectral resolution from the ground has been especially valuable for evincing velocity substructure of the escaping gas motion relative to the exoplanet restframe \citep{2019A&A...629A.110A,2020ApJ...894...97N}, and convincingly associating the signal to an exoplanetary origin as opposed to stellar contamination \citep{2018AJ....156..189C}.  \ion{He}{1} has resulted in at least 14 systems with detections, including \object[GJ 3470]{GJ 3470 b} \citep{2020ApJ...894...97N, 2021A&A...647A.129L}, \object{HD 63433 c} \citep{2022AJ....163...68Z}, WASP 107 b \citep{2019A&A...623A..58A,2020AJ....159..115K}, HD 209458 b \citep{2019A&A...629A.110A}, WASP 69 b \citep{2020AJ....159..278V}, WASP 52 b \citep{2020AJ....159..278V}, HD 189733 b \citep{2021A&A...647A.129L}, and HAT-P-18 b \citep{2021ApJ...909L..10P}.  An observational picture of mass loss in hot Saturns appears to be lacking, at least in part due to their intrinsic rarity.

@@

\subsection{Ohmic dissipation-driven mass loss}
The Ohmic dissipation mechanism made two key observable predictions.  Anomalous heating efficiency, $\epsilon$, should initially increase as a function of equilibrium temperature, then degrade as irradiation exceeds equilibrium temperatures of about 2000 K.  Second, hot Saturns ($\lesssim0.5 M_\mathrm{Jup}$) should undergo runaway evaporation, whereas hot Jupiters ($\gtrsim 1 M_\mathrm{Jup}$) should reach stable equilibrium---albeit inflated---radii after Gyr timescales.  Both of these outcomes predated data that could validate them, and these outcomes differ from the behavior of other heating mechanisms, such as photoevaporation or tides.

@@

\noindent yielding an incredibly short $<$5 Myr timescale for \hatpb, assuming a typical $\epsilon\sim0.01$.  The order of magnitude of this inflation timescale is so fleetingly short that---according to this scenario---\hatpb~must be in the runaway stage of inflation, rapidly losing mass, and growing in surface area to fuel a positive feedback loop.  Under this interpretation, \hatpb~ would represent an example of a new category of planet system that is doomed to evaporate entirely due to the Ohmic dissipation mechanism, as predicted by \citet{2011ApJ...738....1B}.  At Roche lobe overflow, a 5 Myr inflation timescale may imply an instantaneous $\dot{M}_\mathrm{infl}>10^{3}$ M$_\oplus$/Gyr.

@@

The second panel from the top illustrates a countervailing effect: the increase in flux actually triggers a decrease in anomalous heating efficiency, $\epsilon$. The greater stellar energy couples into the planet less effectively than the weaker radiation did, roughly balancing out. Together the effects nearly cancel when computing the inflationary lifetimes, yielding nearly identical curves in the bottom panel: a secularly evolving main sequence history gives almost the same inflation timescale as a rapidly increasing subgiant.  It is important to emphasize that the anomalous heating efficiency found by \citet{2018AJ....155..214T} is agnostic to what the actual heating mechanism is: either XUV Irradiation or Ohmic dissipation both have to obey the trend of $\hat{\epsilon}_G(T_\mathrm{eq})$.

@@

We report in-transit line profile variability, which we attribute to the delicate interplay of planetary and stellar winds.  We identify several avenues for future work, including additional monitoring of the line profile variability to probe the stellar-and-planetary wind interaction.  The large signal should be perceptible in other spectral tracers, such as metal lines in the UV.  We offer a list other planets that may be likely to exhibit mass loss under the Ohmic dissipation and XUV irradiation scenarios.
